---
title: "IntroductionToMarkdown"
author: "NRX07"
date: "11/05/2020"
---

# Heading 1

Use # for main heading.

## Heading 2

Use ## for sub-heading 1.

### Heading 3

Use ### for sub-heading 2.

- Use *here_* to print in *italics*  
- Use **here_** to print in **bold**  
- Use `here_` to print in a `Box`  

```{r setval , echo=TRUE}
library(knitr)
opts_chunk$set(echo=FALSE)#This will set the global option of all the chunks.
```

```{r printing , echo=TRUE}
#We can overwrite global options like this for indivisual chunks.
print("Hello world")#If you are making a plot do not forget to use fig.height and fig.width to set the border size of the plot.  We can set the echo=FALSE id we dont want the code. We can set the results="hide" then we can use the variable `{varnames_}` in normal lines too. Rmd gets converted to md and then it gets converted in html for us to see. we can set the results='asis' to get the output as it is on the html page. For this we can use xtable which a handly function. If the code inside the chunk requires a lot of computational power then we cans et cache=TRUE but we can not set it globally, if the code changes the cache ie, th e output will update too hence we have to see that one code chunk is not dependant on another.
```
xtable (xtable package) Generate a nicely formatted table for LaTeX or a basic table for HTML. x – An R object that could conceivably be formatted into a table.  
Use [name](link) to go to various websites.  
[Youtube](https://www.youtube.com/)  
See rest from [here](https://rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)  

